Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
fix math
huge code listings slain.

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

\begin{figure*}
\lstinputlisting[label=viterbi,caption=Viterbi]{viterbi_simple.hs}
\end{figure*}

@@

We wrote a basic search function in only twenty-two lines of code; as one of 
its parameters we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.

@@

\begin{figure*}
\lstinputlisting[label=search,caption=Search]{search_simple.hs}
\end{figure*}

>>>>

\begin{eqnarray}    
V_{j}^{M}(i) = \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}},\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}},\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{I}(i) = \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + max \left\{
\begin{array}{l l}
V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}},\\
V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{D}(i) = max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}},\\
V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\end{eqnarray}\label{viterbi_eqn}

\begin{eqnarray}    
V_{j}^{\prime M}(i) = e^{\prime}_{M_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}},\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}},\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\nonumber\\
V_{j}^{\prime I}(i) = e^{\prime}_{I_{j}}(x_{i}) + min \left\{
\begin{array}{l l}
V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}},\\
V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}},\\
\end{array} \right.\nonumber\\
V_{j}^{\prime D}(i) = min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}},\\
V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\nonumber\\
\text{where } a^{\prime}_{s} = - \log a_{s} \nonumber \\
\text{and } e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}} \nonumber
\end{eqnarray}\label{viterbi_log_eqn}

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{verbatim}
viterbi s n o
  -- consume an observation AND a node
  | s == mat = let trans t prev = 
                   (score + tProb (n-1) t + (eProb n o), mat:path)
                    where (score, path) = viterbi prev (n - 1) (o - 1)
                in minimum $ [
                    trans m_m mat,
                    trans i_m ins,
                    trans d_m del
                    ]
\end{verbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

In this second approach, we used the 
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

@@

We wrote a basic search function in only twenty-two lines of code.
The central logic of this search function is simply:
\begin{verbatim}
search' (s1:s2:seeds) oldPop hist age =
  let newPop = mutate oldPop
      score = fst $ minimum newPop
      newHist = score : hist
    in if accept newHist age then
          if terminate newHist age then
            (minimum newPop, newHist)
          else
            search' seeds newPop newHist (age + 1)
       else
         if terminate hist age then
           (minimum oldPop, hist)
         else
           search' seeds oldPop hist (age + 1)
\end{verbatim}
As one of the parameters to the top-level \texttt{search} function,
we provide a \texttt{SearchStrategy} data type, which we use to 
define specific search techniques.
We require a \texttt{SearchStrategy} to 
define four functions: \texttt{mutate}, \texttt{accept}, \texttt{terminate}, 
and \texttt{initialize}; we use the Haskell module system to allow sharing of 
some of these functions between different search strategies.
For example, both 
\textit{random mutation hill climb} and \textit{simulated annealing} use the 
same initialize and mutate functions, but very different accept functions.
In 
all, we implemented four search strategies: \textit{random mutation hill 
climb}, \textit{simulated annealing}, \textit{multistart simulated annealing}, 
and \textit{genetic algorithm}.
We found the implementation of the search 
function and search strategies to be pleasant, and the resulting code to be 
simple and elegant.
In an imperative language with object orientation like Ruby 
or Python, we likely would have used a class hierarchy to accomplish this.
We 
feel that this would have required more code, because object orientation is 
meant for encapsulating data as well as functions on those data; here we only 
wanted to encapsulate functions, as the data did not change.

@@
